---
title: "Canopy Height Models"
---

```{r setup, include=FALSE}
source("R/setup_rgl.R")
```

## Relevant Resources

-   [Code](https://github.com/liamirwin/LPS_lidRtutorial/blob/master/R/03_chm.R)
-   [lidRbook section](https://r-lidar.github.io/lidRbook/dsm.html)

## Overview

This code demonstrates the creation of a Canopy Height Model (CHM). Similar to the DTM, a CHM is a rasterized summary of the upper-most surface of the height normalized lidar point cloud. Often this is simply generated as maximum height for each cell. CHMs often serve as fundamental layers for vegetation and biodiversity related applications.

We present different algorithms for generating CHMs and provide options for adjusting resolution and filling empty pixels.

## Environment

```{r clear_warnings, warnings = FALSE, message = FALSE}
# Clear environment
rm(list = ls(globalenv()))

# Load packages
library(lidR)
library(terra)
```

## Data Preprocessing

Once again we load the lidar in to memory with `readLAS`. this time however, we sample points randomly to `decimate` the point cloud and simulate lower density lidar data using `decimate_points()`.

The sampling/return density of your point cloud (particularly those originating from the canopy surface for a CHM) dictates the lowest acceptable spatial resolution.

```{r load_lidar_data}
# Load lidar data and reduce point density
las <- readLAS(files = "data/zrh_norm.laz")

# Density of the data is reduced from 20 points/m² to 10 points/m² for example purposes
las <- decimate_points(las, random(density = 10))
```

``` r
# Visualize the lidar point cloud
plot(las)
```

```{r, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
#| label: fig-chm-initial-las
#| fig-cap: "Normalized lidar point cloud used for generating the Canopy Height Model (CHM)."
# Visualize the lidar point cloud
plot(las, bg = "white")
```

## Point-to-Raster Based Algorithm

First we apply a simple method for generating Canopy Height Models (CHMs). 

Below, the `rasterize_canopy()` function with the `p2r()` algorithm assigns the elevation of the highest point within each 2m grid cell to a corresponding pixel. The resulting CHM is then visualized using the `plot()` function.

```{r}
#| label: fig-chm-p2r-2m
#| fig-cap: "Canopy Height Model (CHM) generated at 2m resolution using the point-to-raster (`p2r`) algorithm."
# Generate the CHM using a simple point-to-raster based algorithm
chm <- rasterize_canopy(las = las, res = 2, algorithm = p2r())

# Visualize the CHM
plot(chm)
```

Next, by increasing the resolution of the CHM to 1m (reducing the grid cell size), we get a more detailed representation of the canopy, but also have more empty pixels.

```{r}
#| label: fig-chm-p2r-1m
#| fig-cap: "CHM generated at a higher resolution of 1m using the `p2r` algorithm, showing more detail and some empty pixels."
# Make spatial resolution 1 m
chm <- rasterize_canopy(las = las, res = 1, algorithm = p2r())
plot(chm)
```

Further increasing to 0.5m causes more empty pixels where no valid points are present. (see tiny white dots on CHM)

```{r}
#| label: fig-chm-p2r-subcircle
#| fig-cap: "A 0.5m resolution CHM created using the `p2r` algorithm with the `subcircle` option to fill gaps."
# Using the 'subcircle' option turns each point into a disc of 8 points with a radius r
chm <- rasterize_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.15))
plot(chm)
```

The `rasterize_canopy()` function with the `p2r()` algorithm allows the use of the `subcircle` option, which turns each lidar point into a disc of 8 points with a specified radius. This can help to capture more fine-grained canopy details, especially from lower density point clouds, resulting in a complete CHM.

Increasing the `subcircle` radius may not necessarily result in meaningful CHMs, as it could lead to over-smoothing or loss of important canopy information.

```{r}
#| label: fig-chm-p2r-large-subcircle
#| fig-cap: "CHM generated with an increased `subcircle` radius, with potential over-smoothing."
# Increasing the subcircle radius, but it may not have meaningful results
chm <- rasterize_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.8))
plot(chm)
```


The `p2r()` algorithm also allows filling empty pixels using TIN (Triangulated Irregular Network) interpolation, which can help in areas with sparse lidar points to obtain a smoother CHM.

```{r}
#| label: fig-chm-p2r-na-fill
#| fig-cap: "CHM with empty pixels filled using Triangulated Irregular Network (TIN) interpolation."
# We can fill empty pixels using TIN interpolation
chm <- rasterize_canopy(las = las, res = 0.5, algorithm = p2r(subcircle = 0.0, na.fill = tin()))
plot(chm)
```

## Triangulation Based Pitfree Algorithm

The `rasterize_canopy` function can also use the Khosravipour et al. 2014 `pitfree()` algorithm with specified height thresholds and a maximum edge length to generate a CHM. This algorithm aims to correct depressions in the CHM surface, especially designed to prevent gaps in CHMs from low density lidar.

::: callout-tip
## Pit-free algorithm

Check out [Khosravipour et al. 2014](https://www.ingentaconnect.com/content/asprs/pers/2014/00000080/00000009/art00003?crawler=true) to see the original implementation of the algorithm!
:::

```{r}
#| label: fig-chm-pitfree
#| fig-cap: "CHM generated using the `pitfree` algorithm by Khosravipour et al. (2014)."
# Using the Khosravipour et al. 2014 pit-free algorithm with specified thresholds and maximum edge length
thresholds <- c(0, 5, 10, 20, 25, 30)
max_edge <- c(0, 1.35)
chm <- rasterize_canopy(las = las, res = 0.5, algorithm = pitfree(thresholds, max_edge))
plot(chm)
```

The `subcircle` option can also be used with the `pitfree()` algorithm to create finer spatial resolution CHMs with subcircles for each lidar point, similar to the point-to-raster based algorithm.

```{r}
#| label: fig-chm-pitfree-subcircle
#| fig-cap: "CHM generated using the `pitfree` algorithm combined with the `subcircle` option for finer detail."
# Using the 'subcircle' option with the pit-free algorithm
chm <- rasterize_canopy(las = las, res = 0.25, algorithm = pitfree(thresholds, max_edge, 0.1))
plot(chm)
```

## Post-Processing

CHMs can be post-processed by smoothing or other manipulations. Here, we demonstrate post-processing using the `terra::focal()` function for average smoothing within a 3x3 moving window.

```{r}
#| label: fig-chm-smoothed
#| fig-cap: "The `pitfree` CHM after post-processing with a 3x3 mean focal filter for smoothing."
# Post-process the CHM using the 'terra' package and focal() function for smoothing
ker <- matrix(1, 3, 3)
schm <- terra::focal(chm, w = ker, fun = mean, na.rm = TRUE)

# Visualize the smoothed CHM
plot(schm)
```

## Exercises

#### E1.

Using the `p2r()` and `pitfree()` (defining your own arguments) create two CHMs with the same resolution and plot them. What differences do you notice?

#### E2.

Using `terra::focal()` with the `w = matrix(1, 5, 5)` and `fun = max`, plot a manipulated CHM using one of the CHMs you previously generated.

#### E3.

Create a 10 m CHM using a algorithm of your choice. Would this information still be useful at this scale?

## Conclusion

This tutorial covered different algorithms for generating Canopy Height Models (CHMs) from lidar data using the `lidR` package in R. It includes point-to-raster-based algorithms and triangulation-based algorithms, as well as post-processing using the `terra` package.
