---
title: "Regions of Interest"
---

```{r setup, include=FALSE}
source("R/setup_rgl.R")
```

## Relevant Resources

-   [Code](https://github.com/liamirwin/LPS_lidRtutorial/blob/master/R/07_s1_roi.R)

-   [lidRbook section](https://r-lidar.github.io/lidRbook/engine.html#engine-clip)

## Overview

We demonstrate the selection of regions of interest (ROIs) from LiDAR data. Geometries like circles and rectangles are selected based on coordinates. Complex geometries are extracted from shapefiles to clip specific areas.

## Environment

```{r clear warnings, warnings = FALSE, message=FALSE}
# Clear environment
rm(list = ls(globalenv()))

# Load packages
library(lidR)
library(sf)
```

## Simple Geometries

### Load LiDAR Data and Inspect

We start by loading some LiDAR data and inspecting its header and number of point records.

```{r load_and_inspect_lidar_data}
las <- readLAS(files = 'data/zrh_norm.laz')
# Inspect the header and the number of point records
las@header
las@header$`Number of point records`
```

### Select Circular and Rectangular Areas

We can select circular and rectangular areas from the LiDAR data based on specified coordinates and radii or dimensions.

```{r select_circular_areas}
# Establish coordinates
x <- 2670592
y <- 1258890

# Select a circular area
circle <- clip_circle(las = las, xcenter = x, ycenter = y, radius = 30)

# Inspect the circular area and the number of point records
circle
circle@header$`Number of point records`
```

``` r
# Plot the circular area
plot(circle)
```

```{r plot_circle, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(circle, bg = "white")
```

We can do the same with a rectangular area by defining corner coordinates.

```{r select_rectangular_areas}
# Select a rectangular area
rect <- clip_rectangle(las = las, xleft = x, ybottom = y, xright = x + 40, ytop = y + 30)
```

``` r
# Plot the rectangular area
plot(rect)
```

```{r plot_rectangle, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(rect, bg = "white")
```

We can also supply multiple coordinate pairs to clip multiple ROIs.

```{r select_multiple}
# Select multiple random circular areas
x <- runif(2, x, x)
y <- runif(2, 1258840, 1258890)

plots <- clip_circle(las = las, xcenter = x, ycenter = y, radius = 10)
```

``` r
# Plot each of the multiple circular areas
plot(plots[[1]])
```

```{r plot_1, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(plots[[1]], bg = "white")
```

``` r
# Plot each of the multiple circular areas
plot(plots[[2]])
```

```{r plot_2, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(plots[[2]], bg = "white")
```

## Extraction of Complex Geometries from Shapefiles

We demonstrate how to extract complex geometries from shapefiles using the `clip_roi()` function from the `lidR` package.

We use the `sf` package to load an ROI and then clip to its extents.

```{r extraction_complex_geometries_sf, warning=FALSE, message=FALSE}
# Load the shapefile using sf
stand_bdy <- sf::st_read(dsn = "data/roi/roi.gpkg", quiet = TRUE)

# Plot the LiDAR header information without the map
plot(las@header, map = FALSE)

# Plot the stand boundary areas on top of the LiDAR header plot
plot(stand_bdy, add = TRUE, col = "#08B5FF39")

# Extract points within the stand boundary using clip_roi()
stand <- clip_roi(las = las, geometry = stand_bdy)
```

``` r
# Plot the extracted points within the planting areas
plot(stand)
```

```{r plot_stand, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Plot the extracted points within the stand area
plot(stand, bg = "white")
```

## Clipping ROIs with a catalog

We clip the `LAS` data in the catalog using specified coordinate groups.

```{r clip_las, warning = FALSE, message = FALSE}

ctg <- catalog("data/ctg_norm")

# Set coordinate groups
x <- c(2670578, 2671234, 2671499, 2671755, 2671122)
y <- c(1258601, 1259050, 1259450, 1259900, 1258750)

# Visualize coordinate groups
plot(ctg)
points(x, y)

# Clip plots
rois <- clip_circle(las = ctg, xcenter = x, ycenter = y, radius = 30)
```

``` r
plot(rois[[1]])
```

```{r visualize_clipped_ctg, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(rois[[1]], bg = "white")
```

``` r
plot(rois[[3]])
```

```{r visualize_clipped_ctg_two, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize the LiDAR data with a default color palette
plot(rois[[3]], bg = "white")
```

## Validate clipped data

We validate the clipped LAS data using the `las_check` function.

```{r validate_clipped}
las_check(rois[[1]])
las_check(rois[[3]])
```

## Independent files (e.g. plots) as catalogs

We read an individual LAS file as a catalog and perform operations on it.

```{r delete_dems, echo = FALSE, results = FALSE}
# Instructions for cleaning up any existing .lax files
# (Note: Please replace 'path' with the appropriate path)
path <- paste0(tempdir())
file_list <- list.files(path)
delete_tif <- file_list[grep("\\.tif$", file_list)]
delete_las <- file_list[grep("\\.laz$", file_list)]
file.remove(file.path(path, delete_tif))
file.remove(file.path(path, delete_las))
```

```{r independent_files, warning = FALSE, message = FALSE}
# Read single file as catalog
ctg <- readLAScatalog(folder = "data/zrh_norm.laz")

# Set options for output files
opt_output_files(ctg) <- paste0(tempdir(),"/{XCENTER}_{XCENTER}")

# Write file as .laz
opt_laz_compression(ctg) <- TRUE

# Get random plot locations and clip
x <- runif(n = 4, min = ctg$Min.X, max = ctg$Max.X)
y <- runif(n = 4, min = ctg$Min.Y, max = ctg$Max.Y)
rois <- clip_circle(las = ctg, xcenter = x, ycenter = y, radius = 10)
```

```{r plots_catalog, warning = FALSE, message = FALSE}
# Read catalog of plots
ctg_plots <- readLAScatalog(tempdir())

# Set independent files option
opt_independent_files(ctg_plots) <- TRUE
opt_output_files(ctg_plots) <- paste0(tempdir(),"/{XCENTER}_{XCENTER}")

# Generate plot-level terrain models
rasterize_terrain(las = ctg_plots, res = 1, algorithm = tin())
```

```{r check_plot}
# Check files
path <- paste0(tempdir())
file_list <- list.files(path, full.names = TRUE)
file <- file_list[grep("\\.tif$", file_list)][[1]]

# plot dtm
plot(terra::rast(file))
```

## Conclusion

This concludes our tutorial on selecting simple geometries and extracting complex geometries from geopackage files (or shapefiles etc...) using the `lidR` package in R.
