---
title: "Individual Tree Detection & Segmentation"
---

```{r setup, include=FALSE}
source("R/setup_rgl.R")
```

## Relevant resources

-   [Code](https://github.com/liamirwin/LPS_lidRtutorial/blob/master/R/08_s2_its.R)
-   [lidRbook section](https://r-lidar.github.io/lidRbook/itd-its.html)

## Overview

This code demonstrates individual tree segmentation (ITS) using LiDAR data. It covers CHM-based and point cloud-based methods for tree detection and segmentation. The code also shows how to extract metrics at the tree level and visualize them.

## Environment

```{r clear warnings, warnings = FALSE, message=FALSE}
# Clear environment
rm(list = ls(globalenv()))

# *Ensure 'concaveman' is installed for tree segmentation*
if (!requireNamespace("concaveman", quietly = TRUE)) {
  install.packages("concaveman")
}

# Load all required packages
library(concaveman)
library(lidR)
library(sf)
library(terra)

# Read in LiDAR file and set some color palettes
las <- readLAS("data/zrh_norm.laz")
col <- height.colors(50)
col1 <- pastel.colors(900)
```

## CHM based methods

We start by creating a Canopy Height Model (CHM) from the LiDAR data. The `rasterize_canopy()` function generates the CHM using a specified resolution (`res`) and a chosen algorithm, here `p2r(0.15)`, to compute the percentiles.

```{r chm_creation}
# Generate CHM
chm <- rasterize_canopy(las = las, res = 0.5, algorithm = p2r(0.15))
plot(chm, col = col)
```

After building the CHM, we visualize it using a color palette (`col`).

## Optionally smooth the CHM

Optionally, we can smooth the CHM using a kernel to remove small-scale variations and enhance larger features like tree canopies.

```{r chm_smoothing}
# Generate kernel and smooth chm
kernel <- matrix(1, 3, 3)
schm <- terra::focal(x = chm, w = kernel, fun = median, na.rm = TRUE)
plot(schm, col = col)
```

Here, we smooth the CHM using a median filter with a 3x3 kernel. The smoothed CHM (`schm`) is visualized using a color palette to represent height values.

## Tree detection

Next, we detect tree tops using the smoothed CHM. The `locate_trees()` function identifies tree tops based on local maxima.

```{r tree_detection, warning = FALSE}
# Detect trees
ttops <- locate_trees(las = schm, algorithm = lmf(ws = 2.5))
ttops
plot(chm, col = col)
plot(ttops, col = "black", add = TRUE, cex = 0.5)
```

The detected tree tops (`ttops`) are plotted on top of the CHM (`chm`) to visualize their positions.

## Segmentation

Now, we perform tree segmentation using the detected tree tops. The `segment_trees()` function segments the trees in the LiDAR point cloud based on the previously detected tree tops.

``` r
# Segment trees using dalponte
las <- segment_trees(las = las, algorithm = dalponte2016(chm = schm, treetops = ttops))
```

```{r tree_segmentation, echo = FALSE, message = FALSE, warning=FALSE}
# Segment trees using dalponte
las <- segment_trees(las = las, algorithm = dalponte2016(chm = schm, treetops = ttops))
```

```{r tree_segmentation_legnth}
# Count number of trees detected and segmented
length(unique(las$treeID) |> na.omit())
```

``` r
# Visualize all trees
plot(las, color = "treeID")
```

```{r visualize_trees, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6, warning = FALSE}
# Visualize using intensity values as colors
plot(las, color = "treeID", bg = "white")
```

```{r}
# Select trees by ID
tree25 <- filter_poi(las = las, treeID == 25)
tree125 <- filter_poi(las = las, treeID == 125)
```

``` r
plot(tree25, size = 4)
```

```{r visualize_tree_25, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
plot(tree25, size = 4, bg = "white")
```

```         
plot(tree125, size = 3)
```

```{r visualize_tree_125, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
plot(tree125, size = 4, bg = "white")
```

After segmentation, we count the number of trees detected and visualize all the trees in the point cloud. We then select two trees (`tree25` and `tree125`) and visualize them individually.

::: callout-note
## Variability and testing

Forests are **highly** variable! This means that some algorithms and parameters will work better than others depending on the data you have. Play around with algorithms and see which works best for your data.
:::

## Working with rasters

The `lidR` package is designed for point clouds, but some functions can be applied to raster data as well. Here, we show how to extract trees from the CHM without using the point cloud directly.

```{r raster_trees}
# Generate rasterized delineation
trees <- dalponte2016(chm = chm, treetops = ttops)() # Notice the parenthesis at the end
trees

plot(trees, col = col1)
plot(ttops, add = TRUE, cex = 0.5)
```

We create tree objects (`trees`) using the `dalponte2016` algorithm with the CHM and tree tops. The resulting objects are visualized alongside the detected tree tops.

# Point-cloud based methods (no CHM)

In this section, we will explore tree detection and segmentation methods that do not require a CHM.

## Tree detection

We begin with tree detection using the local maxima filtering (lmf) algorithm. This approach directly works with the LiDAR point cloud to detect tree tops.

```{r tree_detection_no_chm}
# Detect trees
ttops <- locate_trees(las = las, algorithm = lmf(ws = 3, hmin = 5))
```

``` r
# Visualize
x <- plot(las)
add_treetops3d(x = x, ttops = ttops, radius = 0.5)
```

```{r visualize_tree_tops , echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
# Visualize
x <- plot(las, bg = "white")
add_treetops3d(x = x, ttops = ttops, radius = 0.5)
```

We detect tree tops using the `lmf` algorithm and visualize them in 3D by adding the tree tops to the LiDAR plot.

## Tree segmentation

Next, we perform tree segmentation using the `li2012` algorithm, which directly processes the LiDAR point cloud.

``` r
# Segment using li
las <- segment_trees(las = las, algorithm = li2012())
```

```{r tree_segmentation_li, echo = FALSE, message = FALSE, warning=FALSE}
# Segment using li algorithm - this takes a bit longer
las <- segment_trees(las = las, algorithm = li2012())
```

``` r
plot(las, color = "treeID")
# This algorithm does not seem pertinent for this dataset.
```

```{r visualize_tree_segmented, echo = FALSE, rgl = TRUE, fig.width = 8, fig.height = 6}
plot(las, color = "treeID", bg = "white")
```

The `li2012` algorithm segments the trees in the LiDAR point cloud based on local neighborhood information. However, it may not be optimal for this specific dataset.

# Extraction of metrics

We will extract various metrics from the tree segmentation results.

## Using `crown_metrics()`

The `crown_metrics()` function extracts metrics from the segmented trees using a user-defined function. We use the length of the Z coordinate to obtain the tree height as an example.

```{r crown_metrics_example}
# Generate metrics for each delineated crown
metrics <- crown_metrics(las = las, func = ~list(n = length(Z)))
metrics
plot(metrics["n"], cex = 0.8)
```

We calculate the number of points (`n`) in each tree crown using a user-defined function, and then visualize the results.

## Applying user-defined functions

We can map any user-defined function at the tree level using the `crown_metrics()` function, just like `pixel_metrics()`. Here, we calculate the convex hull area of each tree using a custom function `f()` and then visualize the results.

```{r user_defined_metrics}
# User defined function for area calculation
f <- function(x, y) {
  # Get xy for tree
  coords <- cbind(x, y)
  
  # Convex hull
  ch <- chull(coords)
  
  # Close coordinates
  ch <- c(ch, ch[1])
  ch_coords <- coords[ch, ]
  
  # Generate polygon
  p <- sf::st_polygon(list(ch_coords))
  
  #calculate area
  area <- sf::st_area(p)

  return(list(A = area))
}

# Apply user-defined function
metrics <- crown_metrics(las = las, func = ~f(X, Y))
metrics
plot(metrics["A"], cex = 0.8)
```

::: callout-tip
## 3rd party metric packages

Remember that you can use 3rd party packages like [`lidRmetrics`](https://github.com/ptompalski/lidRmetrics) for crown metrics too!
:::

## Using pre-defined metrics

Some metrics are already recorded, and we can directly calculate them at the tree level using `crown_metrics()`.

```{r pre_defined_metrics}
metrics <- crown_metrics(las = las, func = .stdtreemetrics)
metrics

# Visualize individual metrics
plot(x = metrics["convhull_area"], cex = 0.8)
plot(x = metrics["Z"], cex = 0.8)
```

We calculate tree-level metrics using `.stdtreemetrics` and visualize individual metrics like convex hull area and height.

## Delineating crowns

The `crown_metrics()` function segments trees and extracts metrics at the crown level.

```{r delineate_crowns}
cvx_hulls <- crown_metrics(las = las, func = .stdtreemetrics, geom = 'convex')
cvx_hulls

plot(cvx_hulls)
plot(ttops, add = TRUE, cex = 0.5)

# Visualize individual metrics based on values
plot(x = cvx_hulls["convhull_area"])
plot(x = cvx_hulls["Z"])
```

We use `crown_metrics()` with `.stdtreemetrics` to segment trees and extract metrics based on crown delineation.

## ITD using `LAScatalog`

In this section, we explore Individual Tree Detection (ITD) using the `LAScatalog`. We first configure catalog options for ITD.

```{r set_catalog_options_itd, warning = FALSE, message = FALSE}
# Load catalog
ctg <- catalog('data/ctg_norm')

# Set catalog options
opt_filter(ctg) <- "-drop_z_below 0 -drop_z_above 50"
opt_select(ctg) <- "xyz"
opt_chunk_size(ctg) <- 500
opt_chunk_buffer(ctg) <- 10
opt_progress(ctg) <- TRUE

# Explicitly tell R to use the is.empty function from the lidR package - avoid terra error
is.empty <- lidR::is.empty

# Detect treetops and plot
ttops <- locate_trees(las = ctg, algorithm = lmf(ws = 3, hmin = 10))
chm <- rasterize_canopy(ctg, algorithm = p2r(), res = 1)
plot(chm)
plot(ttops, add = TRUE, cex = 0.1, col = "red")
```

## Conclusion

This concludes the tutorial on various methods for tree detection, segmentation, and extraction of metrics using the `lidR` package in R.
